Manuscript draft

### 2019-07-08 manuscript/manuscript draft.Rmd
updated manuscript draft

*Model specification.* The effects of climate and latitude on primary productivity were analysed using mixed effects models using the package 'lme4' [@bates_fitting_2015] in R v.3.5.1 [@r_core_team_r:_2018]. The effect of each extracted climate variable on each measure of primary productivity was modelled by specifying the climate variable as a fixed effect. Site altitude was also included as a fixed effect. Random effect was stand nested within area. Data from the temperate regions was heavily skewed towards studies from the old-growth forests of the Pacific Northwest. These forests have very high productivity, and so to reduce any bias from over-sampling of this region, the proportion of global forest cover contributed by each Koeppen climate zone was calculated, and the models weighted according these proportions.

Two models were specified: one with the climate variable as a linear term, and one with the climate variable as a polynomial term. AIC values were calculated for the models and used to select the best model. R-squared values were calculated for the best model. In addition, slope values were calculated for the linear models.

@@

*Allocation to carbon fluxes along latitudinal gradients* Variation in allocation to component carbon fluxes along latitudinal gradients was explored for a selection of paired carbon fluxes: GPP:NPP; NPP:ANPP; NPP:BNPP; ANPP:BNPP; ANPP~foliage~:ANPP~woody; and ANPP~foliage~:ANPP~woodystem~. Three approaches were taken. Firstly, for each set of paired fluxes, measurements taken at the same site and plot, and in the same year, were paired together. The ratio of each pair of measurements was calculated. For each set of paired fluxes, the ratios were regressed against latitude and climate variables, using the linear model specified above. Additionally, sites were classified as "boreal", "temperate", or "tropical", using the FAO ecozone classifications from the ForC database. Non-parametric Wilcoxon t-tests were carried out to compare the ratio means between each pair of ecozone classifications.  
In addition, each carbon flux was regressed against latitude, and the models used to generate a series of point estimates along lines of best fit for each flux. For each pair of fluxes, the ratio of each point estimate was calculated and plotted, to gain an approximation of the variation of the ratio of the paired fluxes with latitude.

@@

Latitude is a strong predictor for many of the carbon fluxes, explaining ~65% of variation in GPP (n = 254, p<0.01), 51% in NPP (n = 116, p<0.01) and 50% in ANPP (n = 259, p<0.01).

Analysis of component fluxes indicates that measures of smaller component fluxes do approximately sum to measures of larger fluxes (fig. 3). Modelled estimates of NPP, generated from the sum of ANPP and BNPP~root~, and ANPP, generated from the sum of ANPP~foliage~ and ANPP~woodystem~, fall completely within the confidence intervals of the regressions of field estimates of NPP and ANPP.

@@

*Which climatic variables are the most important drivers behind the latitudinal pattern in primary productivity?* All major carbon fluxes show similar responses to climate variables. Of the climate variables tested, annual wet days, aridity, cloud cover, mean diurnal temperature range and precipitation seasonality were poor explainers of variation in primary productivity, explaining less than 20% of the variation in each of the carbon fluxes.
We found that both measures of mean annual precipitation used were significant (p<0.05) but poor predictors of productivity, explaining, with the exception of R~auto~, at most 37% of variation in carbon flux. Despite being a poor predictor, the results indicate that there is a saturation point at between 3000 and 4000mm annual precipitation, above which productivity no longer increases.  
Looking across all productivity components, the climate variables which explain the highest amount of variation in carbon fluxes are temperature seasonality, mean annual temperature, annual temperature range, and vapour pressure. For measures of woody productivity (ANPP~woody~ and ANPP~woodystem~), potential evapotranspiration and vapour pressure deficit were also strong predictors.

@@

**Mean annual temperature.** All carbon fluxes show a linear increase with temperature.  
**Temperature seasonality.** Productivity is highest where temperature seasonality = 0. GPP, NPP, ANPP, and BNPP~root~ show a polynomial relationship with temperature seasonality. As seasonality increases, these variables show a more rapid decrease in productivity. ANPP~foliage~ and ANPP~woodystem~ show a linear relationship with temperature seasonality.  
**Annual temperature range.** As with temperature seasonality, productivity is highest at annual temperature ranges of 15$^\circ$C or lower. Again, GPP, NPP, ANPP, and BNPP~root~ show a polynomial relationship with temperature seasonality, and ANPP~foliage~ and ANPP~woodystem~ show a linear relationship with temperature seasonality.  
**Vapour pressure.** Productivity increases linearly with vapour pressure for all carbon fluxes, with the exception of ANPP~foliage~, which saturates at around 0.24.

>>>>

*Model specification.* The effects of climate and latitude on primary productivity were analysed using mixed effects models using the package 'lme4' [@bates_fitting_2015] in R v.3.5.1 [@r_core_team_r:_2018]. The effect of each extracted climate variable on each measure of primary productivity was modelled by specifying the climate variable as a fixed effect. Site altitude was also included as a fixed effect. Random effect was stand nested within area. Data from the temperate regions was heavily skewed towards studies from the old-growth forests of the Pacific Northwest. These forests have very high productivity, and so to ensure that results were not unduly influenced by geographic sampling bias, we tried a version of the model where data were weighted according to forested land area within each Koeppen climate zone. Results were similar between the weighted and unweighted model, so, to avoid problems of over-fitting, the weighted model was dropped, and results from this are not presented here.

For each climate variable, two models were specified: one with the climate variable as a linear term, and one with the climate variable as a polynomial term. AIC values were calculated for the models and used to select the best model. R-squared values were calculated for the best model. In addition, slope values were calculated for the linear models.

Because the magnitude of fluxes varies significantly, in order to facilitate comparisons between regression models for each flux, data for each flux was scaled, to give the data a mean of 0 and standard deviation of 1. As each data set was scaled separately, this does not allow for statistical comparisons of slope values, but does assist in visualising the data.

@@

*Allocation to carbon fluxes along latitudinal gradients* Variation in allocation to component carbon fluxes along latitudinal gradients was explored for a selection of paired carbon fluxes: GPP:NPP; NPP:ANPP; NPP:BNPP; ANPP:BNPP; ANPP~foliage~:ANPP~woody~; and ANPP~foliage~:ANPP~woodystem~. Three approaches were taken. Firstly, for each set of paired fluxes, measurements taken at the same site and plot, and in the same year, were paired together. The ratio of each pair of measurements was calculated. For each set of paired fluxes, the ratios were regressed against latitude and climate variables, using the linear model specified above. Additionally, sites were classified as "boreal", "temperate", or "tropical", using the FAO ecozone classifications from the ForC database. Non-parametric Wilcoxon t-tests were carried out to compare the ratio means between each pair of ecozone classifications.  
In addition, each carbon flux was regressed against latitude, and the models used to generate a series of point estimates along lines of best fit for each flux. For each pair of fluxes, the ratio of each point estimate was calculated and plotted, to gain an approximation of the variation of the ratio of the paired fluxes with latitude.

@@

Latitude is a strong predictor for many of the carbon fluxes, explaining 66% of variation in GPP (n = 254, p<0.0001), 51% in NPP (n = 116, p<0.0001) and 50% in ANPP (n = 259, p<0.0001). For all fluxes, their relationship with latitude was best predicted by the linear model.

Analysis of component fluxes indicates that measures of smaller component fluxes do approximately sum to measures of larger fluxes (fig. 3). Modelled estimates of GPP, generated from the sum of NPP and R auto; NPP, generated from the sum of ANPP and BNPP~root~; and ANPP, generated from the sum of ANPP~foliage~ and ANPP~woodystem~, fall completely within the confidence intervals of the regressions of field estimates of GPP, NPP and ANPP respectively.

@@

*Which climatic variables are the most important drivers behind the latitudinal pattern in primary productivity?* All major carbon fluxes show similar responses to climate variables. Of the climate variables tested, annual wet days, aridity, cloud cover, mean diurnal temperature range and precipitation seasonality were poor explainers of variation in primary productivity, explaining less than 20% of the variation in each of the carbon fluxes.
We found that both measures of mean annual precipitation used were significant (p<0.05) but poor predictors of productivity, explaining, with the exception of R~auto~, at most 37% of variation in carbon flux.

Looking across all productivity components, the climate variables which explain the highest amount of variation in carbon fluxes are temperature seasonality, mean annual temperature, annual temperature range, and annual frost days. Vapour pressure deficit came out as a strong predictor for ANPP~woodystem~, and potential evapotranspiration was a good predictor for BNPP~root~ and ANPP~woodystem~.

@@

**Mean annual temperature.** All carbon fluxes show a linear increase with temperature.  
**Temperature seasonality and annual temperature range.** These two climate variables are strongly correlated, and so only temperature seasonality is presented in figure 4. Productivity is highest where temperature seasonality = 0, and at an annual temperature range of 15$^\circ$C or lower. GPP, NPP, ANPP, and R~root~ show a polynomial relationship with both variables, and BNPP~root~ shows a polynomial relationship with annual temperature range. As seasonality increases, these variables show a more rapid decrease in productivity. ANPP~foliage~ and ANPP~woodystem~ show a linear relationship with both variables.  
**Mean annual precipitation.** Despite being a poor predictor of productivity, the results indicate that, for the majority of fluxes, productivity increases with precipitation, up until a saturation point at between 3000 and 4000mm annual precipitation, above which productivity starts to decrease. The notable exception to this is GPP, which shows a significant (p<0.0001) linear increase with precipitation up to measures of nearly 6000mm annually, although the explanatory power of this model was low (R-squared = 0.15).
**Potential evapotranspiration.** All fluxes, with the exception of BNPP~rootfine~ and R~root~ show a strong increase with potential evapotranspiration at values below 1000. Between 1200 and 1700 productivity peaks, and begins to decrease at values above 1800.
